\chapter{Prior Work}

There is considerable prior work into both improving the usability of databases
and into the visualisation of data inside databases. However, not all of this
work provides practical and usable tools. Rather, it outlines potential ideas
for them. Other prior work creates powerful and excellent tools for working
with databases but they require that the user is already knowledgeable in that
field in order to use them.

\section{General Usability}

A number of the prior papers\cite{Date, Jagadish2007} assert that the usability
of a database system is just as important as the performance or features. This
assertion if made based on the fact that the cost of people is far more than
the cost of machines - now more than ever. The papers define usability as
raising the level of abstraction of the user interface. This is done in order
to allow more people to be able to use the software or to help current users
more easily use the software.

The \emph{Database Usability} paper\cite{Date} outlines some plans and ideas
for improving database usability. For example, text editors, voice inputs and
outputs, and pictorial interfaces. I don't believe that all of these interfaces
would be effective or practical. However, the text editor interface is an
interesting idea and is similar to my project. I also believe that the
pictorial interface may be something that I can use and implement.

The second paper, \emph{Making Database Systems Usable}, notices that improving
usability is the key to improving user engagement. However, this assertion is
made after observing the improvements in search engine quality over the past
few years. It goes on to mention how the problem of querying a database is far
more complex when applied to databases compared to search engines. Many of the
current systems require bespoke software to reach the same level of usability.

\section{Improving Existing Solutions}

One of the papers\cite{Stolte2010} claims that databases that are easy-to-use
generate and improve story-telling, debate, and conversations. Additionally it
provides an interesting concept. It introduces a more effective query language
that can be compiled to SQL in order to be compatible with many different
database systems by speaking their language directly. This solution is similar
to part of my project that I will discuss later.

\section{Different Kinds of Visualisation}

One method of improving usability is providing different methods of
visualisation. The rest of the
papers\cite{Hu2008,Haritsa2010,Yang2010,Jin2010,Chan2009} are introducing
different kinds of visualisation for databases.

\emph{QueryScope}\cite{Hu2008}, \emph{GBLENDER}\cite{Jin2010}, and
\emph{Picasso}\cite{Haritsa2010} provide a method of visualising the queries
that are passed to the database. This allows for similar query patterns to be
identified and, if required, optimised. This is a method of improving usability
that I did not want to pursue as it does not improve the usability for all
users of database management systems. This method only improves the usability
for those that enter queries directly such as database administrators.

Other projects\cite{Yang2010,Chan2009} improve the usability of these systems
by allowing the data in the database to be viewed in different manners. This is
the type of usability improvement that I am interested in working on as it
improves the system for all users. One of the systems\cite{Yang2010} allows for
the generation of various kinds of graphs and visual representations to be
created from the data. The other solution, \emph{Vispedia}\cite{Chan2009},
allows for the querying and viewing of semi-structured date in a structured
way. The example that is given in the paper is \emph{Wikipedia}. Using the
system they have designed they are able to, for instance, ask for the
population of Russia and it will look up the information even though the data
is not strictly structured. This concept is very interesting however I will
explore a different method in my project.
